---
title: 'Practical Machine Learning Course Project - Weight Lifting Exercise Classification'
author: "Henrique Souza (github.com/htssouza)"
output:
  html_document: default
  pdf_document:
    pandoc_args:
    - +RTS
    - -K64m
    - -RTS
---

# Executive Summary

This report describes the process behind the creation of a Machine Learning Model
used to classify weight lifting exercise (Unilateral Dumbbell Biceps Curling) in classes:

. A: exactly according to the specification;
. B: throwing the elbows to the front;
. C: lifting the dumbbell only halfway; 
. D: lowering the dumbbell only halfway; 
. E: throwing the hips to the front (Class E).

More about the research and data used can be found on the following website:
http://groupware.les.inf.puc-rio.br/har#weight_lifting_exercises#ixzz6CzLP0YxO

<strong>TODO: resultado do modelo</strong>

# Exploratory Data Analysis

Our data source urls:

```{r echo = TRUE}
TRAINING_SOURCE_FILE_URL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
TESTING_SOURCE_FILE_URL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
```

```{r echo = FALSE, message = FALSE}
# libraries and data download
library(caret)
library(dplyr)
library(ggplot2)

WORKING_DIRECTORY <- getwd()
TRAINING_FILE_PATH <- file.path(WORKING_DIRECTORY, "pml-training.csv")
TESTING_FILE_PATH <- file.path(WORKING_DIRECTORY, "pml-testing.csv")

if (! file.exists(TRAINING_FILE_PATH)) {
    download.file(url = TRAINING_SOURCE_FILE_URL, destfile = TRAINING_FILE_PATH, method = "curl")
}
if (! file.exists(TESTING_FILE_PATH)) {
    download.file(url = TESTING_SOURCE_FILE_URL, destfile = TESTING_FILE_PATH, method = "curl")
}
```

Loading and splitting the data (training, validating and testing):

```{r echo = TRUE}
NA_STRINGS <- c("NA","#DIV/0!")
training <- read.csv(TRAINING_FILE_PATH, na.strings = NA_STRINGS)
testing <- read.csv(TESTING_FILE_PATH, na.strings = NA_STRINGS)
in.training <- createDataPartition(y = training$class, p = 0.7, list = FALSE)
validating <- training[-in.training, ]
training <- training[in.training, ]
```

How our training dataset looks like:

Rows for each "classe":

Columns with high NA counts.

Outliers.

# Models

Remove unecessary columns

NZV

Outlier removal.

Normalization?

Models

# Model selection

# Accuracy and Residual Analsysis

# Prediction

# Conclusion

# Appendix

Boxplot for each numeric variable per classe:

```{r echo = FALSE, message = FALSE}
numeric.columns <- names(training)[as.character(sapply(training, function(x) { class(x) })) %in% c("numeric", "integer")]
unwanted.columns <- c("X", "raw_timestamp_part_1", "raw_timestamp_part_2", "cvtd_timestamp", "num_window")
wanted.columns <- numeric.columns[! numeric.columns %in%unwanted.columns ]
OUTLIER_IQR_FACTOR <- 2
for(column.name in wanted.columns) {
    df <- select(training, classe, var = !!column.name)
    df <- filter(df, !is.na(var))
    print(ggplot(data = df, aes(y = var, fill = classe)) +
              geom_boxplot(coef = OUTLIER_IQR_FACTOR,
                           outlier.colour = "red",
                           outlier.shape = 16,
                           outlier.size = 2,
                           notch = FALSE) +
              ggtitle(column.name))
}

```

