---
title: 'Practical Machine Learning Course Project - Weight Lifting Exercise Classification'
author: "Henrique Souza (github.com/htssouza)"
output:
  html_document: default
  pdf_document:
    pandoc_args:
    - +RTS
    - -K64m
    - -RTS
---

# Executive Summary

This report describes the process behind the creation of a Machine Learning Model
used to classify weight lifting exercise (unilateral dumbbell biceps curling) in classes:

  * A: exactly according to the specification;
  * B: throwing the elbows to the front;
  * C: lifting the dumbbell only halfway; 
  * D: lowering the dumbbell only halfway; 
  * E: throwing the hips to the front.

More about the research and data used can be found on the following website:
http://groupware.les.inf.puc-rio.br/har#weight_lifting_exercises#ixzz6CzLP0YxO

<strong>TODO: RESULTADO DO MODELO</strong>

# Exploratory Data Analysis

Our data source urls:

```{r echo = TRUE}
TRAINING_SOURCE_FILE_URL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
TESTING_SOURCE_FILE_URL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
```

```{r echo = FALSE, message = FALSE}
# libraries and data download
library(caret)
library(dplyr)
library(ggplot2)

set.seed(12345)

WORKING_DIRECTORY <- getwd()
TRAINING_FILE_PATH <- file.path(WORKING_DIRECTORY, "pml-training.csv")
TESTING_FILE_PATH <- file.path(WORKING_DIRECTORY, "pml-testing.csv")

if (! file.exists(TRAINING_FILE_PATH)) {
    download.file(url = TRAINING_SOURCE_FILE_URL, destfile = TRAINING_FILE_PATH, method = "curl")
}
if (! file.exists(TESTING_FILE_PATH)) {
    download.file(url = TESTING_SOURCE_FILE_URL, destfile = TESTING_FILE_PATH, method = "curl")
}
```

Loading and splitting the data (training, validating and testing):

```{r echo = TRUE}
NA_STRINGS <- c("NA","#DIV/0!")
training <- read.csv(TRAINING_FILE_PATH, na.strings = NA_STRINGS)
testing <- read.csv(TESTING_FILE_PATH, na.strings = NA_STRINGS)
in.training <- createDataPartition(y = training$class, p = 0.7, list = FALSE)
validating <- training[-in.training, ]
training <- training[in.training, ]
```

How our training dataset looks like:

```{r echo = TRUE}
dim(training)
print(table(training$class))
```

Checking the presence of NAs per variable:

```{r echo = FALSE, message = FALSE}
na.per.column <- sapply(training, function(x) sum(is.na(x)) / length(x))
na.per.column <- as.numeric(na.per.column[order(-na.per.column)])
na.stats <- table(cut(na.per.column, 20))
na.stats <- na.stats[na.stats > 0]
```

```{r echo = TRUE}
na.stats
```

For a large number of variables, they have 95% of more of NAs values.
These variables will be ignored on our models.

```{r echo = FALSE, message = FALSE}
unwanted.columns <- c("X", "raw_timestamp_part_1", "raw_timestamp_part_2", "cvtd_timestamp", "new_window", "num_window", "user_name")

na.per.column <- sapply(training, function(x) sum(is.na(x)) / length(x))
na.per.column <- na.per.column[order(-na.per.column)]
almost.empty.columns <- names(na.per.column[as.numeric(na.per.column) >= 0.95])

training <- training %>% 
  select(-unwanted.columns) %>% 
  select(-almost.empty.columns) 
```

Outliers.

You can see more details about the training dataset on the Appendix.

# Models

Remove unecessary columns

NZV

Outlier removal.

Normalization?

Models

# Model selection

# Accuracy and Residual Analsysis

# Prediction

# Conclusion

# Appendix

Variables there are being ignored:

```{r echo = TRUE}
unwanted.columns
almost.empty.columns
```

With the removal of variables with too many NAs, there are no more NearZeroVars as well:

```{r echo = TRUE}
nzv <- nearZeroVar(training, saveMetrics = TRUE)
print(nzv[nzv$nzv,])
```

Boxplot for each numeric variable per classe:

```{r echo = FALSE, message = FALSE}
numeric.columns <- names(training)[as.character(sapply(training, function(x) { class(x) })) %in% c("numeric", "integer")]
OUTLIER_IQR_FACTOR <- 2
for(column.name in numeric.columns) {
    df <- select(training, classe, var = !!column.name)
    df <- filter(df, !is.na(var))
    print(ggplot(data = df, aes(y = var, fill = classe)) +
              geom_boxplot(coef = OUTLIER_IQR_FACTOR,
                           outlier.colour = "red",
                           outlier.shape = 16,
                           outlier.size = 2,
                           notch = FALSE) +
              ggtitle(column.name))
}
```

